Seismic parameter estimation and the Canadian crust

### 2013-12-09 paper/bptectono.tex
tyhor

\documentclass{elsart}
\usepackage{graphicx,multirow,color,natbib,amssymb}
\journal{Tectonophysics}

\graphicspath{ {figures/} }
\DeclareGraphicsExtensions{.png}

@@

\title{Stroboscopic Doppler tomography of FO Aqr}
%% \author[South]{T.R. Marsh\corauthref{cor}}
%% \corauth[cor]{Corresponding author.}
%% \ead{trm@astro.soton.ac.uk}
%% \author[Oxford]{S.R. Duck\thanksref{now}}
%% \thanks[now]{Present address: Systems Engineering and Assessment Ltd., Beckington Castle, PO Box 800, Bath BA3 6TB, UK.}
%% \ead{srd@sea.co.uk}
%% \address[South]{University of Southampton, Department of Physics, Highfield, Southampton SO17 1BJ, UK}
%% \address[Oxford]{University of Oxford, Department of Physics, Nuclear Physics Laboratory, Keble Road, Oxford, OX1 3RH, UK}

%% ------------------------------------------------------------------------ %%
\begin{abstract}

\end{abstract}

@@

The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior form part of the assemblage that comprises the cratonic core of Laurentia [Hammer et al., 2010]. Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield [Hoffman, 1988, Figure 1]. The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and sourthern Superior is composed of roughly eastwest trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6Ga [Thurston et al., 1991]. In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth [O'Neil et al., 2011]. To the east lies the younger, 1.2-0.98Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision [Eaton et al., 2006]. North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga [Thompson et al., 2010]. The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc [Kusky, 1989]. To the west of the Superior are the extensive Interior Platforms comprising the southern extents of the Trans-Hudson Orogen and Hearne Domain as well as the Alberta Orogen. The Interior platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism [Hammer et al., 2010]. Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean to Paleoproterozoic crystalline basement. Between the middle of the Jurasic and Paleocene intense metamorphic, plutonic and uplifting events occured as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coastal mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock [Clowes et al, 1995]. Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale décollement beneath interleaved and duplexed crust [Hammer et al., 2010]. The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through Lithoprobe, a now complete national geoscience project spearheaded by active-source seismic profiling [Clowes et al., 1999]. The full suite of Lithoprobe seismic profiles form a near-continuous transect from the Pacific to the Atlantic Oceans, and have enabled many insights into crustal structure and genesis. Recent summary studies utilizing Lithoprobe data include the analysis of Cook et al [2009] who argue that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration. In another recent Lithoprobe study, Hammer and Clowes [2011] discuss broad structural characteristics of the Canadian crust such as their observation that Achean cratons and their boundaries contain structures very similar to those found in Proterozoic and Phanerozoic orogens. Another key observation is the lack of evidence for significant crustal disruption from magmatic instrusion related to subduction, post-orogenic extension or orogenic collapse. The authors also remark upon the striking uniformity in Moho depth that is imaged between 33km and 43km depth along 20,000 km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. and Hammer and Clowes raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary. The Lith5.0 revised crustal model presented by Perry, Eaton and Forte [2002] was developed from crustal thickness estimates taken from Lithoprobe seismic profiles. This work updates the global Crust5.1 model [Mooney et al, 1998] and introduces broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick (>42.5km) Phanerozoic crust in the south and west of the continental craton.

In this study, we examine the additional constraints that analyses of teleseismic receiver functions place upon Canadian bulk crustal properties when combined with available active source results. It is widely appreciated that receiver functions are sensitive primarily to crustal thickness and Poisson's ratio (or equivalently $V_P/V_S$) (Zhu and Kanomori, 20xx), and that the latter quantity provides some ability to discriminate among possible bulk compositions [Zandt and Ammon 1995]. The constraint on Poisson's ratio and the greater averaging performed by low frequency (0.04-3 Hz) scattered teleseismic waves versus high frequency (10-40 Hz) {\it P}-waves employed in active source studies render the two techniques somewhat complementary. Our data set comprises the majority of teleseismic data available for the Canadian landmass including both permanent and temporary deployments. Many of the temporary deployments are described in [Bostock, Eaton, Snyder]. Three previous studies employing similar methodologies but with more regional emphases deserve special mention: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen [Eaton et al., 2005], a structural and compositional characterization of the Superior Province [Darbyshire et al. 2006] and a crustal evolution analysis of the Hudson Bay region [Thompson et al., 2010]. Results from these studies will serve as benchmarks for our expanded effort. In the following sections we describe the data and methods employed to interrogate the Canadian crust. We proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three studies mentioned above for identical seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the resulting crustal property estimates.  First presented are the average crustal discontinuity and structural patterns across Canada. Second, we take a tour through those geological provinces for which sufficient data exists, focusing on regional variation and characterization. We conclude with a discussion on the implications for our understanding of the composition and evolution of the Canadian Crust.

\subsection{Geological and tectonic summary}
\begin{itemize}

\item Large scale overview/summary of regions under study
\item Justification for selecting these regions?

\end{itemize}

\subsection{Previous geophysical studies}
 \begin{itemize}

\item Lithoprobe and attendent papers
\item Teleseismic studies

\end{itemize}
%% -----------------------------------------------%%

\section{Data and methods}
Our analysis of bulk Canadian continental crust is based on estimates of crustal properties that can be accessed via seismic techniques, namely seismic wave velocities (or velocity ratio) and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source is teleseismic receiver function (hereafter RF) analysis. The remaining two datasets are Crust 1.0, a widely published statistical average of similar regions with a global two degree resolution, and a compilation of pre-processed active source data (Mooney, 2012). Both Crust 1.0 and the active source compilation are used to qualify and compare with the primary dataset.

\subsection{Teleseismic Data Set}
The primary data utilized in this study are computed from teleseismic {\it P} wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between 2000 and 2012 and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, POLARIS, FedNor and Chasme. Seismograms are included for analysis if they fall within a 30 to 100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with sufficiently high signal to noise ratios to permit visual observation of the dominant {\it P} wave energy and with sufficiently impulsive first arrivals to allow arrival onsets to be accurately measured. After selection and filtering more than 80,000 events are available for further processing.

The first stage of processing requires the transformation of teleseismic data into receiver functions. In general terms, this transformation involves deconvolving an approximation of the earthquake source from horizontal recordings of ground motion [Langston, 1979]. The resulting waveforms contain discrete pulses corresponding to {\it S} waves scattered from subsurface discontinuities including the base of the crust, or Moho. Confident identification of these peaks is required to estimate crustal parameters, and improved results can be obtained by transforming the displacement seismogram into their {\it P} and {\it S} contributions based on a 1-D model of propagation. This procedure is accomplished by first rotating the N and E coordinates into radial and transverse directions and then performing a wave field decomposition using the radial and vertical channels [Kennet, 1991]. The direct {\it P} arrival of the signal on the resulting {\it P} wave component is used as an approximation to the source function as {\it P} component impulse response for teleseismic P approximates a delta function. This windowed source estimate is deconvolved from the {\it S} wave component computed from the wave field decomposition.

@@

% Deconvolution equations
\begin{equation}
  r(t) = {\cal F}^{-1} \left[ G(\omega) \right] = {\cal F}^{-1}
 \left[ \frac {\sum_n^N S_n(\omega)P_n^*(\omega)} {\sum_n^N P_n(\omega)P_n^*(\omega) + \delta} \right ],
\end{equation}

@@

A well tested and widely published method for extracting the ratio of {\it P} velocity $V_P$ to {\it S} velocity $V_S$, herafter referred to as $R$, and crustal thickness (or depth to Moho) $H$, is outlined by Zhu and Kanamori [2000], hereafter ZK. This method exploits the dependence of these parameters to the differential arrival times between the {\it S} wave reflected phases $Ps$, $PpPs$, $PsPs$, $PpSs$ and $PsSs$ and the direct {\it P} wave arrival $Pp$ (Fig \ref{fig:reflectedPhases}). Note that $PsPs$ and $PpSs$ are kinematic analogues, such that the energy for these two phases arrive simultaneously for a 1-D model. For a range of slowness values, $p$, the differential arrival times, $t(p)$, trace moveout curves for each phase arrival given by

% Travel time equations
\begin{equation} \label{eq:tps}
t_{Ps}(p_n) = \frac{H}{V_P} \left[ \sqrt{ R^2 - V_P^2p_n^2} - \sqrt{1 - V_P^2p_n^2} \right]
\end{equation}

\begin{equation}
t_{Pps}(p_n) = \frac{H}{V_P} \left[ \sqrt{ R^2 - V_P^2p_n^2} + \sqrt{1 - V_P^2p_n^2} \right]
\end{equation}

@@

RFs are stacked along trial moveout curves for a range of candidate models of $R$ and $H$ to generate a fitness function $s(H,R)$.  Within the stacking procedure, each phase is assigned a weight to account for the expected relative amplitudes of the phases for typical velocity contrasts, with the direct conversion $Ps$ usually characterized by the highest quality signal followed by $PpPs$ and $PpSs$. We chose weights of $w1 = 0.5$, $w2 = 0.3$ , $w3 = -0.2$ for the $Ps$, $PpPs$ and $PpSs$ phases, respectively. A negative weight for the combined $PpSs$ and $PsPs$ phases is required as the polarity of the signal is reversed. Semblance weighting [Eaton, 2006] is employed to reduce the effect of spurious large amplitude noise in the data. The semblance function assigns a weight between zero (incoherent noise) and one (coherent signal). The stacking function is therefore defined as

@@

\begin{equation}
S_m(H,R) = \frac {\left[ \sum_{n=1}^N r_n(t_m) \right]^2}
                 { \sum_{n=1}^N r_n^2(t_m) }
\end{equation}

@@

%% [[\subsection{Full Parameter Method}]]
A limitation of the ZK approach is the requirement of an initial $V_P$ estimate. In principal we may compute the stacking function Eq. \ref{eq:stack}, as a function of all three seismic parameters, $s(H,R,V_P)$. Taking advantage of modern multi-core systems with reasonable quantities of data (fifty to a few hundred RFs) and a moderate search space ($n^3$ parameter candidates where $n \approx 150$) the computational time makes this method easily scalable to hundreds of stations. This ``fullgrid'' approach is analyzed to determine the viability of recovering $V_P$.

The trade-off between $V_P$ and $H$ (and other parameter combinations) and error resulting from data quality are quantified for both the full parameter search and ZK methods by bootstrap resampling [Efron and Tibshirani, 1986]. An estimate for the error is calculated in a bootstrap resampling approach by taking the standard deviation of 1024 parameter estimates each produced with randomly selected RFs, with replacement.

@@

The Crust 1.0 model [Laske et. al., 2013] employs many of these same active source data to provide regular, complete coverage across all of Canada. The model includes estimates of $V_P$ and $V_S$ information with depth at $1^\circ$ resolution in latitude and longitude. In addition to employing local active source data, Crust 1.0 is also constrained by statistically aggregated data from geographically distant but geologically similar regions. It therefore affords a somewhat independent dataset suited for comparison with the crustal estimates calculated in this study.

@@

Three published studies have provided parameter estimates for $R$ and $H$ for subsets of the seismic stations considered here. The most recent study [Thompson et. al., 2010] focused on the northern Canadian Shield and employs both the linear stacking approach used by ZK as well as a phase weighted method. A 2007 study of the Superior Province [Darbyshire et. al, 2007] and a study of the Grenville Orogen [Eaton et. al., 2006] utilize the ZK + semblance approach outlined in section \ref{section:VpVsMethod}.  Comparisons between crustal estimates of $R$ for particular seismic stations are made for those values with corresponding standard errors of less than $\pm$0.06. This error threshold is chosen empirically as that which best partitions those receiver function stacks with visually recognizable energy along phase moveout curves from those where lack of data or poor quality make these moveout curves difficult to distinguish. For each study under comparison, data are reprocessed using the $V_P$ chosen by the study authors. For the 2006 and 2007 studies the ZK + semblance method is used whereas for the Thompson et. al study the linear stacking ZK approach is employed. Both correlation coefficient and mean difference are provided in (Table \ref{table:comparison}) as measures of similarity.

\begin{table}
  \begin{tabular}{ l l l l l l l}
    \cline{4-7}
    & & & \multicolumn{2}{ c }{Correlation Coeff.} & \multicolumn{2}{ c }{RMS difference} \\
    \hline
    Study Authors             & Region         & Num. of Stns &  $H$ & $R$  &$H$(km)& $R$ \\
    \hline
    Thompson et. al. (2010)   & Northern Canadian Sheild & 29 & 0.97 & 0.70 & 0.78 & 0.017 \\
    Darbyshire et. al. (2007) & Superior Province        & 10 & 0.95 & 0.43 & 1.71 & 0.037 \\
    Eaton et. al. (2006)      & Grenville Orogen         & 26 & 0.90 & 0.61 & 1.03 & 0.033 \\
    \hline
  \end{tabular}
  \caption{Comparison of $R$ and $H$ estimates with three published studies}
\label{table:comparison}

@@

All three studies reported $H$ estimates that correlate strongly with the crustal thickness parameters computed here. The data from the northern Canadian Shield shows strong correlation in both $H$ and $R$. The lower correlation observed in estimates of $R$ than $H$ is not unexpected as the range of $R$ is relatively narrow and small changes ($\approx 2\%$) are physically meaningful. A corollary of this is greater sensitivity of $R$ to differences in procedure such as the choice of deconvolution algorithm and selection and scaling of data. The results from Thompson et. al. [2010] also corrispond most closely to the parameter estimates computed here, with the mean difference for both parameters being less than half of the averaged bootstrap error computed in this study. Results published by {\it Eaton et. al.} [2006] show reasonable correlations in both $R$ and $H$ and a mean difference for both parameters that is roughly equal to the averaged uncertainty in the data. Parameter estimates from the Superior Province [Darbyshire et. al., 2007] show the lowest correlation in $R$ as well as the highest mean difference in both parameters. These differences are well over the averaged computed error, and can not be easily explained by deviation related to noise in the data.

@@

We now proceed to compare our results computed using the ZK approach with those from the full grid search (Table \ref{table:ZKvsFG}). We note that the velocity ratio $R$ has a correlation between the two datasets of 0.76 whereas $H$ has a correlation of 0.51. The low correlation between crustal thickness estimates can be explained by the strong tradeoff between $H$ and $V_P$ through the quantity

@@

%% -----------------------------------------------%%
\section{Conclusions}

@@

\section{bibliography}

[[ADD GRANITE DIORITE REFERENCE]]

>>>>

\documentclass[review]{elsarticle}
\usepackage{graphicx,multirow,color,natbib,amssymb}
\journal{Tectonophysics}

\graphicspath{ {figures/} }

@@

\title{Seismic parameter estimation and the Canadian crust}

\author[ubc]{B. Postlethwaite\corref{cor1}}
\ead{bpostlet@eos.ubc.ca}
\author[ubc]{M. G. Bostock}
\ead{bostock@eos.ubc.ca}
\author[ubc]{N. I. Christensen}
\author[gsc]{D. B. Snyder}

\cortext[cor2]{Corresponding author}

\address[ubc]{Earth and Ocean Sciences, UBC, Vancouver BC, Canada}
\address[gsc]{Geological Survey Canada, Natural Resources Canada, Ottawa ON, Canada}

%% ------------------------------------------------------------------------ %%
\begin{abstract}
It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. Existing models based on fractionating mantle composition or evolving mantle convection require accurate estimates of the geological and geophysical properties of crustal provinces to better understand the details of early continental formation (R. Durrheim, W. Mooney, 1991). Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of 1-D structure. Additionally, we investigate the feasibility of employing a variant on this method which allows resolution of $V_S$ and $V_P$ uniquely. Bootstrap error analysis is performed for each station dataset and subsets of these results are compared to the results for matching stations from previously published studies. Cross-analyzing these results with the mineral and rock seismic property database of Christensen (1996) will afford improved constraints on bulk geological composition of the Canadian landmass. These results will be used to make structural characterizations of the crust as well as evaluate competing models of crustal formation.
\end{abstract}

@@

The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior form part of the assemblage that comprises the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and sourthern Superior is composed of roughly eastwest trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011). To the east lies the younger, 1.2-0.98Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989). To the west of the Superior are the extensive Interior Platforms comprising the southern extents of the Trans-Hudson Orogen and Hearne Domain as well as the Alberta Orogen. The Interior platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean to Paleoproterozoic crystalline basement. Between the middle of the Jurasic and Paleocene intense metamorphic, plutonic and uplifting events occured as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coastal mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale décollement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through Lithoprobe, a now complete national geoscience project spearheaded by active-source seismic profiling (Clowes et al., 1999). The full suite of Lithoprobe seismic profiles form a near-continuous transect from the Pacific to the Atlantic Oceans, and have enabled many insights into crustal structure and genesis. Recent summary studies utilizing Lithoprobe data include the analysis of Eaton (2005) who examines seismic evidence for several hypotheses of the origin of continental Moho: relic Moho, underplating, metamorphic front and regional décollement. Seismic evidence of preserved island arc Moho in Cascadia supports the first hypothesis while the presence of high-velocity lower-crustal xenoliths in southern Alberta suggests thick mafic underplating and evidence for the second. Eclogite mineral assemblages forming a deep crustal root below the Grenville Front lends support to the third hypothesis, that at least in some places, continental Moho forms by large scale eclogitization of the lower crust. Finally, seismic imaging of subhorizontal reflections seperated by dipping reflection fabrics beneath the southern Canadian Cordillera suggests decoupling of the upper crust from the mantle during deformation and is taken as evidence for the fourth hypothesis, that an overprinted Moho may serve as a subhorizontal regional dècollement. Continuing from this effort, a study by Cook et al. (2009) argues that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration. In another recent Lithoprobe study, Hammer and Clowes (2011) discuss broad structural characteristics of the Canadian crust such as their observation that Achean cratons and their boundaries contain structures very similar to those found in Proterozoic and Phanerozoic orogens. Another key observation is the lack of evidence for significant crustal disruption from magmatic instrusion related to subduction, post-orogenic extension or orogenic collapse. The authors also remark upon the striking uniformity in Moho depth that is imaged between 33km and 43km depth along 20,000 km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. and Hammer and Clowes raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary. The Lith5.0 revised crustal model presented by Perry, Eaton and Forte (2002) was developed from crustal thickness estimates taken from Lithoprobe seismic profiles. This work updates the global Crust5.1 model (Mooney et al, 1998) and introduces broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick (>42.5km) Phanerozoic crust in the south and west of the continental craton.

In this study, we examine the additional constraints that analyses of teleseismic receiver functions place upon Canadian bulk crustal properties when combined with available active source results. It is widely appreciated that receiver functions are sensitive primarily to crustal thickness and Poisson's ratio (or equivalently $V_P/V_S$) (Zhu and Kanomori, 2000), and that the latter quantity provides some ability to discriminate among possible bulk compositions (Zandt and Ammon, 1995). The constraint on Poisson's ratio and the greater averaging performed by low frequency (0.04-3 Hz) scattered teleseismic waves versus high frequency (10-40 Hz) {\it P}-waves employed in active source studies render the two techniques somewhat complementary. Our data set comprises the majority of teleseismic data available for the Canadian landmass including both permanent and temporary deployments. Many of the temporary deployments are described in (Bostock et al, 2010). Three previous studies employing similar methodologies but with more regional emphases deserve special mention: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen (Eaton et al., 2005), a structural and compositional characterization of the Superior Province (Darbyshire et al. 2006) and a crustal evolution analysis of the Hudson Bay region (Thompson et al., 2010). Results from these studies will serve as benchmarks for our expanded effort. In the following sections we describe the data and methods employed to interrogate the Canadian crust. We proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three studies mentioned above for identical seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the resulting crustal property estimates.  First presented are the average crustal discontinuity and structural patterns across Canada. Second, we take a tour through those geological provinces for which sufficient data exists, focusing on regional variation and characterization. We conclude with a discussion on the implications for our understanding of the composition and evolution of the Canadian Crust.

\section{Data and methods}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{stationMap.pdf}
  \caption{Map showing major Canadian geological provinces and seismic stations contributing to the data discussed in this study.}
  \label{map:stationMap}
\end{figure}

Our analysis of bulk Canadian continental crust is based on estimates of crustal properties that can be accessed via seismic techniques, namely seismic wave velocities (or velocity ratio) and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source is teleseismic receiver function (hereafter RF) analysis. The remaining two datasets are Crust 1.0, a widely published statistical average of similar regions with a global two degree resolution, and a compilation of pre-processed active source data (Mooney, 2012). Both Crust 1.0 and the active source compilation are used to qualify and compare with the primary dataset.

\subsection{Teleseismic Data Set}
The primary data utilized in this study are computed from teleseismic {\it P} wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between 2000 and 2012 and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, POLARIS, FedNor and Chasme. Seismograms are included for analysis if they fall within a 30 to 100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with sufficiently high signal to noise ratios to permit visual observation of the dominant {\it P} wave energy and with sufficiently impulsive first arrivals to allow arrival onsets to be accurately measured. After selection and filtering more than 80,000 events are available for further processing. Seismic stations contributing to the data utilized in this study are shown in figure \ref{map:stationMap}.

The first stage of processing requires the transformation of teleseismic data into receiver functions. In general terms, this transformation involves deconvolving an approximation of the earthquake source from horizontal recordings of ground motion (Langston, 1979). The resulting waveforms contain discrete pulses corresponding to {\it S} waves scattered from subsurface discontinuities including the base of the crust, or Moho. Confident identification of these peaks is required to estimate crustal parameters, and improved results can be obtained by transforming the displacement seismogram into their {\it P} and {\it S} contributions based on a 1-D model of propagation. This procedure is accomplished by first rotating the N and E coordinates into radial and transverse directions and then performing a wave field decomposition using the radial and vertical channels (Kennet, 1991). The direct {\it P} arrival of the signal on the resulting {\it P} wave component is used as an approximation to the source function as {\it P} component impulse response for teleseismic P approximates a delta function. This windowed source estimate is deconvolved from the {\it S} wave component computed from the wave field decomposition.

@@

% Deconvolution equations
\begin{equation}
  r(t) = {\cal F}^{-1} \left( G(\omega) \right) = {\cal F}^{-1}
 \left[ \frac {\sum_n^N S_n(\omega)P_n^*(\omega)} {\sum_n^N P_n(\omega)P_n^*(\omega) + \delta} \right ],
\end{equation}

@@

A well tested and widely published method for extracting the ratio of {\it P} velocity $V_P$ to {\it S} velocity $V_S$, herafter referred to as $R$, and crustal thickness (or depth to Moho) $H$, is outlined by Zhu and Kanamori (2000), hereafter ZK. This method exploits the dependence of these parameters to the differential arrival times between the {\it S} wave reflected phases $Ps$, $PpPs$, $PsPs$, $PpSs$ and $PsSs$ and the direct {\it P} wave arrival $Pp$ (Fig \ref{fig:reflectedPhases}). Note that $PsPs$ and $PpSs$ are kinematic analogues, such that the energy for these two phases arrive simultaneously for a 1-D model. For a range of slowness values, $p$, the differential arrival times, $t(p)$, trace moveout curves for each phase arrival given by

% Travel time equations
\begin{equation} \label{eq:tps}
t_{Ps}(p_n) = \frac{H}{V_P} \left( \sqrt{ R^2 - V_P^2p_n^2} - \sqrt{1 - V_P^2p_n^2} \right)
\end{equation}

\begin{equation}
t_{Pps}(p_n) = \frac{H}{V_P} \left( \sqrt{ R^2 - V_P^2p_n^2} + \sqrt{1 - V_P^2p_n^2} \right)
\end{equation}

@@

RFs are stacked along trial moveout curves for a range of candidate models of $R$ and $H$ to generate a fitness function $s(H,R)$.  Within the stacking procedure, each phase is assigned a weight to account for the expected relative amplitudes of the phases for typical velocity contrasts, with the direct conversion $Ps$ usually characterized by the highest quality signal followed by $PpPs$ and $PpSs$. We chose weights of $w1 = 0.5$, $w2 = 0.3$ , $w3 = -0.2$ for the $Ps$, $PpPs$ and $PpSs$ phases, respectively. A negative weight for the combined $PpSs$ and $PsPs$ phases is required as the polarity of the signal is reversed. Semblance weighting (Eaton, 2006) is employed to reduce the effect of spurious large amplitude noise in the data. The semblance function assigns a weight between zero (incoherent noise) and one (coherent signal). The stacking function is therefore defined as

@@

\begin{equation}
S_m(H,R) = \frac {\left( \sum_{n=1}^N r_n(t_m) \right)^2}
                 { \sum_{n=1}^N r_n^2(t_m) }
\end{equation}

@@

A limitation of the ZK approach is the requirement of an initial $V_P$ estimate. In principal we may compute the stacking function Eq. \ref{eq:stack}, as a function of all three seismic parameters, $s(H,R,V_P)$. Taking advantage of modern multi-core systems with reasonable quantities of data (fifty to a few hundred RFs) and a moderate search space ($n^3$ parameter candidates where $n \approx 150$) the computational time makes this method easily scalable to hundreds of stations. This ``fullgrid'' approach is analyzed to determine the viability of recovering $V_P$.

The trade-off between $V_P$ and $H$ (and other parameter combinations) and error resulting from data quality are quantified for both the full parameter search and ZK methods by bootstrap resampling (Efron and Tibshirani, 1986). An estimate for the error is calculated in a bootstrap resampling approach by taking the standard deviation of 1024 parameter estimates each produced with randomly selected RFs, with replacement.

@@

The Crust 1.0 model (Laske et. al., 2013) employs many of these same active source data to provide regular, complete coverage across all of Canada. The model includes estimates of $V_P$ and $V_S$ information with depth at $1^\circ$ resolution in latitude and longitude. In addition to employing local active source data, Crust 1.0 is also constrained by statistically aggregated data from geographically distant but geologically similar regions. It therefore affords a somewhat independent dataset suited for comparison with the crustal estimates calculated in this study.

@@

Three published studies have provided parameter estimates for $R$ and $H$ for subsets of the seismic stations considered here. The most recent study (Thompson et. al., 2010) focused on the northern Canadian Shield and employs both the linear stacking approach used by ZK as well as a phase weighted method. A 2007 study of the Superior Province (Darbyshire et. al, 2007) and a study of the Grenville Orogen (Eaton et. al., 2006) utilize the ZK + semblance approach outlined in section \ref{section:VpVsMethod}.  Comparisons between crustal estimates of $R$ for particular seismic stations are made for those values with corresponding standard errors of less than $\pm$0.06. This error threshold is chosen empirically as that which best partitions those receiver function stacks with visually recognizable energy along phase moveout curves from those where lack of data or poor quality make these moveout curves difficult to distinguish. For each study under comparison, data are reprocessed using the $V_P$ chosen by the study authors. For the 2006 and 2007 studies the ZK + semblance method is used whereas for the Thompson et. al study the linear stacking ZK approach is employed. Both correlation coefficient and mean difference are provided in (Table \ref{table:comparison}) as measures of similarity.

\begin{table}
  \begin{tabular}{ p{3cm} p{2cm} p{1cm} p{0.5cm} p{0.5cm} p{0.5cm} p{0.5cm}}
    & & & \multicolumn{2}{ c }{Corr. Coeff.} & \multicolumn{2}{ c }{RMS difference} \\
    \hline
    Study Authors & Region & Num. Stns &  $H$ & $R$  &$H$ (km)& $R$ \\
    \hline
    Thompson et. al. (2010) & N. Canadian Shield & 29 & 0.97 & 0.70 & 0.78 & 0.017 \\
    Darbyshire et. al. (2007) & Superior Province & 10 & 0.95 & 0.43 & 1.71 & 0.037 \\
    Eaton et. al. (2006) & Grenville Orogen & 26 & 0.90 & 0.61 & 1.03 & 0.033 \\
    \hline
  \end{tabular}
  \caption{Comparison of $R$ and $H$ estimates with three published studies}
\label{table:comparison}

@@

All three studies reported $H$ estimates that correlate strongly with the crustal thickness parameters computed here. The data from the northern Canadian Shield shows strong correlation in both $H$ and $R$. The lower correlation observed in estimates of $R$ than $H$ is not unexpected as the range of $R$ is relatively narrow and small changes ($\approx 2\%$) are physically meaningful. A corollary of this is greater sensitivity of $R$ to differences in procedure such as the choice of deconvolution algorithm and selection and scaling of data. The results from Thompson et. al. (2010) also corrispond most closely to the parameter estimates computed here, with the mean difference for both parameters being less than half of the averaged bootstrap error computed in this study. Results published by {\it Eaton et. al.} (2006) show reasonable correlations in both $R$ and $H$ and a mean difference for both parameters that is roughly equal to the averaged uncertainty in the data. Parameter estimates from the Superior Province (Darbyshire et. al., 2007) show the lowest correlation in $R$ as well as the highest mean difference in both parameters. These differences are well over the averaged computed error, and can not be easily explained by deviation related to noise in the data.

@@

We now proceed to compare our results computed using the ZK approach with those from the full grid search. We note that the velocity ratio $R$ has a correlation between the two datasets of 0.76 whereas $H$ has a correlation of 0.51. The low correlation between crustal thickness estimates can be explained by the strong tradeoff between $H$ and $V_P$ through the quantity

@@

@@

\section{References}
